---
title: "Getting Started with the PRIME Ecosystem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the PRIME Ecosystem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The **PRIME ecosystem** consists of three interconnected tools for the analysis of
transcription initiation data (e.g., CAGE-seq):

| Tool | Type | Purpose |
|---|---|---|
| [PRIMEprep](https://github.com/anderssonlab/PRIMEprep) | Shell pipeline | Raw FASTQ → QC → trimming → mapping → BigWig |
| [PRIME](https://github.com/anderssonlab/PRIME) | R package | CTSS quantification, divergent loci, normalization, noise estimation |
| [PRIMEmodel](https://github.com/anderssonlab/PRIMEmodel) | R + Python | Genome-wide prediction of regulatory elements |

The tools form a linear pipeline, and **CAGEfightR** provides core CTSS data
structures that PRIME extends:

```
FASTQ files
    │
    ▼
┌──────────┐
│ PRIMEprep │  Shell: QC → trim → rRNA filter → STAR map → G-correction → BigWig
└────┬─────┘
     │ BigWig files (plus/minus strand)
     ▼
┌──────────┐
│  PRIME   │  R package (extends CAGEfightR)
│          │  ● quantify CTSSs (via CAGEfightR)
│          │  ● pool replicates, subsample, normalize
│          │  ● call tag clusters & divergent loci
│          │  ● estimate noise
│          │  ● complexity & saturation analysis
└────┬─────┘
     │ CTSS RangedSummarizedExperiment objects
     ▼
┌────────────┐
│ PRIMEmodel │  R + Python (LightGBM)
│            │  ● genome-wide prediction of regulatory elements
│            │  ● focal prediction on defined regions
│            │  ● post-processing & BED output
└────────────┘
```

The **dependency chain** is: CAGEfightR → PRIME → PRIMEmodel.

## Installation

### 1. Bioconductor dependencies (CAGEfightR)

```{r install-bioc}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("CAGEfightR")
```

### 2. PRIME

```{r install-prime}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("anderssonlab/PRIME")
```

### 3. PRIMEmodel (requires Python >= 3.9)

```{r install-primemodel}
devtools::install_github("anderssonlab/PRIMEmodel")
```

PRIMEmodel requires a Python environment (>= 3.9) with LightGBM and supporting
packages. See the detailed installation guide at
<https://github.com/anderssonlab/PRIMEmodel/blob/main/install_PRIMEmodel.md>.

### 4. PRIMEprep

PRIMEprep is a shell-based preprocessing pipeline. Clone the repository:

```bash
git clone https://github.com/anderssonlab/PRIMEprep.git
```

PRIMEprep depends on the following external tools (recommended versions):

| Tool | Version |
|---|---|
| STAR | v2.7.3a |
| fastp | v0.23.4 |
| samtools | v1.20.0 |
| rRNAdust | v1.02 |
| bedGraphToBigWig | v4.0 |
| preseq | v2.0 |
| FastQC | v0.12.1 |
| bedtools | v2.31.0 |
| Perl | v5.38.0 |
| openjdk | v20.0.0 |

## Quick start

Load the key libraries:

```{r libraries}
library(CAGEfightR)
library(PRIME)
```

The PRIME package includes small example data for testing:

```{r example-data}
# Load example design matrix
dir_design_matrix <- system.file("extdata", "design_matrix_k562_10pct.tsv",
                                 package = "PRIME")
design_matrix <- read.table(dir_design_matrix, header = TRUE, sep = "\t")
rownames(design_matrix) <- design_matrix$Name

# Load example BigWig files
dir_bw <- system.file("extdata", "cage_bw", package = "PRIME")
bw_plus  <- rtracklayer::BigWigFileList(file.path(dir_bw, design_matrix$BigWigPlus))
bw_minus <- rtracklayer::BigWigFileList(file.path(dir_bw, design_matrix$BigWigMinus))
names(bw_plus) <- names(bw_minus) <- design_matrix$Name

# Quantify CTSSs
ctss <- CAGEfightR::quantifyCTSSs(plusStrand = bw_plus,
                                  minusStrand = bw_minus,
                                  design = design_matrix)
```

## See also

- `vignette("02-preprocessing")` — preprocessing CAGE data with PRIMEprep
- `vignette("03-ctss-processing")` — CTSS quantification and QC with PRIME
- `vignette("08-end-to-end-workflow")` — complete pipeline walkthrough
- [PRIMEprep repository](https://github.com/anderssonlab/PRIMEprep)
- [PRIMEmodel repository](https://github.com/anderssonlab/PRIMEmodel)
- [Paper analysis code](https://github.com/anderssonlab/nucCAGE_PRIME_paper)
