---
title: "Calling Divergent Loci"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calling Divergent Loci}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Divergent loci are genomic regions with divergent transcription: plus and
minus strand transcription initiation events that originate close together and
point away from each other. They are a hallmark of active enhancers and promoters.

PRIME provides two methods for calling divergent loci from CTSS data:

1. **Summit-based** (`PRIME::divergentLociTCsSummit()`) — identifies summit
   positions of tag clusters on each strand and pairs them into divergent loci.
   This is the recommended native PRIME approach.

2. **FANTOM5-style / TC-based** (`PRIME::divergentLoci()`) — first calls
   unidirectional tag clusters with CAGEfightR and then pairs them into
   divergent loci. This mirrors the approach used in FANTOM5.

## Setup

```{r libraries}
library(CAGEfightR)
library(PRIME)
```

Assumes a CTSS `RangedSummarizedExperiment` object `ctss` is available; see
`vignette("03-ctss-processing")` for how to build one.

## Method 1: Summit-based divergent loci (PRIME native)

This approach directly uses CTSS counts to identify divergent loci via summit
detection. 

```{r summit-method}
DLs <- PRIME::divergentLociTCsSummit(
        ctss         = ctss,
        callingAssay = "counts.noSingletons"
    )
```

For large datasets it is efficient to process per chromosome:

```{r summit-method}
CTSSs_by_chr <- split(ctss, GenomicRanges::seqnames(ctss))

DLs <- lapply(names(CTSSs_by_chr), function(chr) {
    PRIME::divergentLociTCsSummit(
        ctss         = CTSSs_by_chr[[chr]],
        callingAssay = "counts.noSingletons"
    )
})

# Combine results across chromosomes
DLs <- do.call(c, DLs)
```

## Method 2: FANTOM5-style divergent loci

This approach first calls unidirectional tag clusters and then pairs them:

```{r fantom-method}
# Pool CTSSs and call unidirectional tag clusters
ctss_pooled <- CAGEfightR::calcPooled(ctss, inputAssay = "counts.noSingletons")
ctss_pooled <- subset(ctss_pooled, score > 0)
TCs <- CAGEfightR::clusterUnidirectionally(ctss_pooled)

# Call divergent loci
DLs_fantom <- PRIME::divergentLoci(
    object = TCs,
    ctss   = ctss_pooled
)
```

## Quantifying divergent loci

Once divergent loci have been called, expression can be quantified across
all samples using:

```{r quantify}
DLs_quant <- PRIME::quantifyDivergentLoci(
    object       = DLs,
    ctss         = ctss,
    inputAssay   = "counts",
    outputColumn = "score"
)
```

## See also

- `vignette("03-ctss-processing")` — producing the CTSS object
- `vignette("06-noise-estimation")` — estimating divergent background noise
- `vignette("07-prediction")` — predicting enhancers and promoters with PRIMEmodel
- [Paper analysis code — Divergent Loci](https://github.com/anderssonlab/nucCAGE_PRIME_paper/tree/main/3.DivergentLoci)
