---
title: "Normalization, Subsampling, and Batch Handling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Normalization, Subsampling, and Batch Handling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette describes how PRIME handles normalization across libraries,
subsampling for saturation analyses, and batch effects in multi-experiment
CTSS datasets.

Assumes a CTSS `RangedSummarizedExperiment` object `ctss` is available;
see `vignette("03-ctss-processing")` for how to build one.

## Setup

```{r libraries}
library(PRIME)
library(SummarizedExperiment)
```

## Size factor normalization

PRIME provides two size-factor-based normalization functions:

```{r size-factor}
# Compute size factors (e.g., from total tags or an external method)
size_factors <- colData(ctss)$totalTags

# Normalize raw counts by size factors
ctss_norm <- PRIME::normalizeBySizeFactors(
    ctss,
    sizeFactors  = size_factors,
    inputAssay   = "counts",
    outputAssay  = "normalized"
)

# Normalize to TPM scaled by size factors
ctss_norm <- PRIME::TPMnormalizeBySizeFactors(
    ctss,
    sizeFactors  = size_factors,
    inputAssay   = "counts",
    outputAssay  = "normalizedTPM"
)
```

## Subsampling

Subsampling is used to equalize library depth before comparative analyses and
for sequencing saturation experiments.

### Subsample to an absolute target depth

```{r subsample-target}
ctss_sub <- PRIME::subsampleTarget(ctss, target = 30e6)
```

### Subsample to a proportion of the library

```{r subsample-prop}
ctss_prop <- PRIME::subsampleProportion(ctss, proportion = 0.5)
```

## Batch handling

When data are generated across multiple experimental batches, PRIME provides
functions to assess and correct for batch effects:

```{r batch}
# Assess support across batches
# (batch is a factor vector with one entry per sample)
batch_support <- PRIME::calcBatchSupport(
    ctss,
    batch        = colData(ctss)$Batch,
    inputAssay   = "counts",
    unexpressed  = 0
)

# Conditional normalization (normalize within batches)
ctss_cond <- PRIME::conditionalNormalize(
    ctss,
    batch        = colData(ctss)$Batch,
    inputAssay   = "counts",
    outputAssay  = "normalizedBatch"
)
```

## Export normalized BigWig files

After normalization, export per-sample BigWig files for visualization or
downstream tools:

```{r write-bw}
PRIME::writeBw(
    ctss,
    replicates = colnames(ctss),
    dir        = "bw_output/"
)
```

## See also

- `vignette("03-ctss-processing")` — CTSS quantification and basic QC
- `vignette("06-noise-estimation")` — estimating genomic background noise
- `vignette("08-end-to-end-workflow")` — complete pipeline walkthrough
- [Paper analysis code](https://github.com/anderssonlab/nucCAGE_PRIME_paper)
