---
title: "End-to-End Workflow: From FASTQ to Regulatory Elements"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-End Workflow: From FASTQ to Regulatory Elements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette walks through the complete PRIME ecosystem pipeline, from raw
CAGE sequencing reads to a genome-wide map of predicted regulatory elements.
The pipeline involves three tools:

```
FASTQ files
    │
    ▼
┌──────────┐
│ PRIMEprep │  Shell: QC → trim → rRNA filter → STAR map → G-correction → BigWig
└────┬─────┘
     │ BigWig files (plus/minus strand)
     ▼
┌──────────┐
│  PRIME   │  R package (extends CAGEfightR)
│          │  ● quantify CTSSs (via CAGEfightR)
│          │  ● pool replicates, subsample, normalize
│          │  ● call tag clusters & divergent loci
│          │  ● estimate noise
│          │  ● complexity & saturation analysis
└────┬─────┘
     │ CTSS RangedSummarizedExperiment objects
     ▼
┌────────────┐
│ PRIMEmodel │  R + Python (LightGBM)
│            │  ● genome-wide prediction of regulatory elements
│            │  ● focal prediction on defined regions
│            │  ● post-processing & BED output
└────────────┘
```

Each step is described in detail in the dedicated vignettes:

- `vignette("02-preprocessing")` — PRIMEprep
- `vignette("03-ctss-processing")` — CTSS quantification and QC
- `vignette("04-divergent-loci")` — divergent loci calling
- `vignette("05-normalization-batches")` — normalization and batch handling
- `vignette("06-noise-estimation")` — genomic background noise
- `vignette("07-prediction")` — regulatory element prediction

This vignette provides a concise end-to-end example.

## Step 1: Preprocess CAGE data with PRIMEprep

Run the PRIMEprep shell pipeline on each FASTQ file:

```bash
cd PRIMEprep
./PRIMEprep.sh \
    -f /data/sample1.fastq.gz \
    -g /ref/genome.fa \
    -b /ref/STAR_index \
    -d /ref/rRNAdust_db \
    -t 16 \
    -o /results/sample1
```

The key output is `bw_files/` containing strand-specific BigWig files:

```
/results/sample1/bw_files/sample1.plus.bw
/results/sample1/bw_files/sample1.minus.bw
```

See `vignette("02-preprocessing")` for full details on PRIMEprep.

## Step 2: Quantify CTSSs with CAGEfightR and PRIME

```{r setup-libs}
library(CAGEfightR)
library(PRIME)
library(SummarizedExperiment)
library(GenomeInfoDb)
```

### Load the design matrix

```{r design}
design <- read.table("design_matrix.tsv", header = TRUE, sep = "\t")
rownames(design) <- design$Name
# Required columns: Name, BigWigPlus, BigWigMinus
# Optional columns: CellLine, Type, Replicate, Batch, ...
```

### Build the CTSS object

```{r ctss}
bw_plus  <- rtracklayer::BigWigFileList(design$BigWigPlus)
bw_minus <- rtracklayer::BigWigFileList(design$BigWigMinus)
names(bw_plus) <- names(bw_minus) <- design$Name

ctss <- CAGEfightR::quantifyCTSSs(plusStrand = bw_plus,
                                  minusStrand = bw_minus,
                                  design      = design)
ctss <- CAGEfightR::calcTotalTags(ctss)
ctss <- CAGEfightR::calcTPM(ctss)
ctss <- CAGEfightR::calcPooled(ctss)

# Restrict to standard chromosomes
ctss <- GenomeInfoDb::keepSeqlevels(ctss,
                                    GenomeInfoDb::seqlevels(ctss)[1:23],
                                    pruning.mode = "coarse")
```

### Pool replicates and subsample

```{r pool-sub}
# Pool biological replicates
ctss_pooled <- PRIME::poolReplicates(
    ctss,
    replicates = paste(colData(ctss)$CellLine, colData(ctss)$Type, sep = "_")
)
ctss_pooled <- CAGEfightR::calcTPM(ctss_pooled)

# Subsample to the minimum library depth for fair comparison
min_depth <- min(colData(ctss)$totalTags)
ctss_sub  <- PRIME::subsampleTarget(ctss, target = min_depth)
ctss_sub  <- CAGEfightR::calcTPM(ctss_sub)

# Remove singletons
ctss_clean <- PRIME::rmSingletons(ctss_sub)
```

## Step 3: Call divergent loci and assess complexity

### Call divergent loci

```{r divergent-loci}
CTSSs_by_chr <- split(ctss_clean, GenomicRanges::seqnames(ctss_clean))

DLs <- lapply(names(CTSSs_by_chr), function(chr) {
    PRIME::divergentLociTCsSummit(
        ctss         = CTSSs_by_chr[[chr]],
        callingAssay = "counts"
    )
})
DLs <- do.call(c, DLs)
```

### Assess sequencing saturation

```{r complexity}
complexity <- PRIME::calcCTSSComplexity(ctss,
                                        step           = 2e6,
                                        minCTSSsupport = 1,
                                        CTSSunexpressed = 0)
```

### Estimate genomic background noise

```{r noise}
# Requires mask (blacklist + hotspots + exons) and mappable regions
# See vignette("06-noise-estimation") for full preparation steps
noise <- PRIME::estimateNoise(
    ctss_clean,
    mask       = mask,
    mappable   = mappable,
    strand     = "*",
    inputAssay = "counts",
    quantiles  = seq(0, 1, 0.0001)
)
```

## Step 4: Predict regulatory elements with PRIMEmodel

```{r primemodel}
library(PRIMEmodel)
library(reticulate)

# ctss_clean is the CTSS object from Steps 2-3
result <- PRIMEmodel::predict(
    ctss_rse        = ctss_clean,
    python_path     = reticulate::py_config()$python,
    score_threshold = 0.75,
    score_diff      = 0.1,
    num_cores       = NULL,
    keep_tmp        = FALSE
)
```

Alternatively, run focal prediction on the divergent loci identified in Step 3:

```{r focal}
result_focal <- PRIMEmodel::predictFocal(
    ctss_rse    = ctss_clean,
    regions     = DLs,
    python_path = reticulate::py_config()$python
)
```

## Full reproducibility

The complete analysis code for the associated publication (Einarsson,
Navamajiti, et al. 2026) is available at:
<https://github.com/anderssonlab/nucCAGE_PRIME_paper>

That repository contains step-by-step R Markdown notebooks for all analyses
described in the paper.

## See also

- `vignette("01-getting-started")` — installation of all tools
- `vignette("02-preprocessing")` — PRIMEprep in depth
- `vignette("03-ctss-processing")` — CTSS quantification and QC
- `vignette("04-divergent-loci")` — divergent loci calling
- `vignette("05-normalization-batches")` — normalization and batch handling
- `vignette("06-noise-estimation")` — noise estimation
- `vignette("07-prediction")` — PRIMEmodel prediction in depth
- [PRIMEprep repository](https://github.com/anderssonlab/PRIMEprep)
- [PRIMEmodel repository](https://github.com/anderssonlab/PRIMEmodel)
- [Paper analysis code](https://github.com/anderssonlab/nucCAGE_PRIME_paper)
