---
title: "CTSS Quantification and Quality Control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CTSS Quantification and Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette covers the quantification of CAGE transcription start sites (CTSSs)
from BigWig files produced by PRIMEprep, followed by quality control,
replicate pooling, subsampling, and complexity analysis.

CTSSs are single-base-pair positions with associated 5' end read counts. The
`CAGEfightR::quantifyCTSSs()` function reads strand-specific BigWig files and
produces a `RangedSummarizedExperiment` object that serves as the primary data
structure throughout the PRIME workflow.

The PRIME package includes small example data for testing basic functionality.

## Setup

```{r libraries}
library(CAGEfightR)
library(rtracklayer)
library(PRIME)
library(SummarizedExperiment)
library(GenomeInfoDb)
library(ggplot2)
```

## Design matrix

The design matrix is a tab-separated file with one row per sample. Required
columns are `Name`, `BigWigPlus`, and `BigWigMinus`. Additional columns (e.g.,
`CellLine`, `Type`, `Replicate`) can be used for downstream grouping.

Using the bundled example data:

```{r design-example}
dir_design <- system.file("extdata", "design_matrix_first10pct.tsv",
                          package = "PRIME")
design <- read.table(dir_design, header = TRUE, sep = "\t")
rownames(design) <- design$Name
```

## Load BigWig files

Using the bundled example data:

```{r load-bw-example}
dir_bw <- system.file("extdata", "cage_bw", package = "PRIME")
bw_plus  <- rtracklayer::BigWigFileList(file.path(dir_bw, design$BigWigPlus))
bw_minus <- rtracklayer::BigWigFileList(file.path(dir_bw, design$BigWigMinus))
names(bw_plus) <- names(bw_minus) <- design$Name
```

## Quantify CTSSs

```{r quantify}
ctss <- CAGEfightR::quantifyCTSSs(plusStrand = bw_plus,
                                  minusStrand = bw_minus,
                                  design = design)
ctss <- CAGEfightR::calcTotalTags(ctss)
ctss <- CAGEfightR::calcTPM(ctss)
ctss <- CAGEfightR::calcPooled(ctss)
```

Inspect the object:

```{r inspect}
ctss
head(assay(ctss, "counts"))
colData(ctss)
```

## Keep standard chromosomes

For large datasets it is recommended to restrict to the standard chromosomes
before further processing:

```{r seqlevels}
ctss <- GenomeInfoDb::keepStandardChromosomes(ctss, pruning.mode = "coarse")
```

## Complexity and saturation analysis

PRIME provides several functions for assessing library complexity and sequencing
saturation. These are particularly useful for deciding whether additional
sequencing depth is needed.

```{r complexity}
# CTSS-level complexity: how many unique CTSS positions are detected?
complexity <- PRIME::calcCTSSComplexity(ctss,
                                        step         = 2e6,
                                        minCTSSsupport = 1,
                                        CTSSunexpressed  = 0)

# Gene-level complexity (requires a TxDb object)
# gene_complexity <- PRIME::calcGeneComplexity(ctss, txModels = txdb, step = 2e6)

# Divergent loci complexity (requires called divergent loci)
# dl_complexity <- PRIME::calcDivergentLociComplexity(ctss, loci = dl, step = 2e6)
```

```{r complexity-plot}
# Set up the dataframe for plotting
df_complexity <- data.frame(
  totalTags   = complexity$totalTags,
  numberCTSSs = complexity$numberCTSSs,
  sample      = complexity$sample
)

# Remove the missing points
df_complexity <- na.omit(df_complexity)

# Plot
ggplot(df_complexity, aes(x = totalTags,
                          y = numberCTSSs,
                          color = sample,
                          group = sample)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_point(size = 1.2) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Total Tags", y = "Number of CTSSs", color = NULL,
       title = "Library complexity saturation curves") +
  theme_bw()
```

## Pool biological replicates

`PRIME::poolReplicates()` sums counts across replicates to produce a pooled
CTSS object. The `replicates` argument should be a character vector of group
labels (one per column in `ctss`).

```{r pool}
ctss_pooled <- PRIME::poolReplicates(
    ctss,
    replicates = paste(colData(ctss)$CellLine,
                       colData(ctss)$Type,
                       sep = "_")
)
ctss_pooled <- CAGEfightR::calcTPM(ctss_pooled)
```

## Subsample to common sequencing depth

Subsampling equalizes library sizes before comparative analyses.

```{r subsample}
min_depth <- min(colData(ctss)$totalTags)
ctss_sub  <- PRIME::subsampleTarget(ctss, target = min_depth)
ctss_sub  <- CAGEfightR::calcTPM(ctss_sub)
```

## Remove singletons

Singletons (CTSSs observed only once across all samples) are typically removed
to reduce noise before calling tag clusters or divergent loci.

```{r rm-singletons}
ctss_clean <- PRIME::rmSingletons(ctss_sub)
```

## See also

- `vignette("getting-started")` — installation and overview
- `vignette("preprocessing")` — generating BigWig files with PRIMEprep
- `vignette("divergent-loci")` — calling divergent loci from CTSSs
- `vignette("normalization-batches")` — normalization and batch handling
- [Paper analysis code](https://github.com/anderssonlab/nucCAGE_PRIME_paper)
