---
title: "End-to-End Workflow: From FASTQ to Regulatory Elements"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-End Workflow: From FASTQ to Regulatory Elements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette walks through the complete PRIME toolkit pipeline, from raw
CAGE sequencing reads to a genome-wide map of predicted regulatory elements.
The pipeline involves three tools:

```
FASTQ files
    │
    ▼
┌──────────┐
│ PRIMEprep │  Shell: QC → trim → rRNA filter → STAR map → G-correction → BigWig
└────┬─────┘
     │ BigWig files (plus/minus strand)
     ▼
┌──────────┐
│  PRIME   │  R package (extends CAGEfightR)
│          │  ● quantify CTSSs (via CAGEfightR)
│          │  ● pool replicates, subsample, normalize
│          │  ● call tag clusters & divergent loci
│          │  ● estimate noise
│          │  ● complexity & saturation analysis
└────┬─────┘
     │ CTSS RangedSummarizedExperiment objects
     ▼
┌────────────┐
│ PRIMEmodel │  R + Python (LightGBM)
│            │  ● genome-wide prediction of regulatory elements
│            │  ● focal prediction on defined regions
│            │  ● post-processing & BED output
└────────────┘
```

Each step is described in detail in the dedicated vignettes:

- `vignette("preprocessing")` — PRIMEprep
- `vignette("ctss-processing")` — CTSS quantification and QC
- `vignette("divergent-loci")` — divergent loci calling
- `vignette("normalization-batches")` — normalization and subsampling
- `vignette("noise-estimation")` — genomic background noise
- `vignette("prediction")` — regulatory element prediction

This vignette provides a concise end-to-end example.

## Step 1: Preprocess CAGE data with PRIMEprep

Run the PRIMEprep shell pipeline on each FASTQ file:

```bash
cd PRIMEprep
./PRIMEprep.sh \
    -f /data/sample1.fastq.gz \
    -g /ref/genome.fa \
    -b /ref/STAR_index \
    -d /ref/rRNAdust_db \
    -t 16 \
    -o /results/sample1
```

The key output is `bw_files/` containing strand-specific BigWig files:

```
/results/sample1/bw_files/sample1.plus.bw
/results/sample1/bw_files/sample1.minus.bw
```

See `vignette("preprocessing")` for full details on PRIMEprep.

## Step 2: Quantify CTSSs with CAGEfightR and PRIME

```{r setup-libs}
library(CAGEfightR)
library(PRIME)
library(SummarizedExperiment)
library(GenomeInfoDb)
library(ggplot2)
```

### Bundled example data path

```{r data-paths}
dir_design <- system.file("extdata", "design_matrix_first10pct.tsv",
                          package = "PRIME")
dir_bw <- system.file("extdata", "cage_bw", package = "PRIME")
```

### Load the design matrix using :

```{r load-bw}
design <- read.table(dir_design, header = TRUE, sep = "\t")
rownames(design) <- design$Name

# Required columns: Name, BigWigPlus, BigWigMinus
# Optional columns: CellLine, Type, Replicate, Batch, ...
```

### Build the CTSS object

```{r ctss}
bw_plus  <- rtracklayer::BigWigFileList(file.path(dir_bw, design$BigWigPlus))
bw_minus <- rtracklayer::BigWigFileList(file.path(dir_bw, design$BigWigMinus))
names(bw_plus) <- names(bw_minus) <- design$Name

ctss <- CAGEfightR::quantifyCTSSs(plusStrand = bw_plus,
                                  minusStrand = bw_minus,
                                  design      = design)
ctss <- CAGEfightR::calcTotalTags(ctss)
ctss <- CAGEfightR::calcTPM(ctss)
ctss <- CAGEfightR::calcPooled(ctss)

# Restrict to standard chromosomes
ctss <- GenomeInfoDb::keepStandardChromosomes(ctss, pruning.mode = "coarse")
```

### Pool replicates and subsample

```{r pool-sub}
# Pool biological replicates
ctss_pooled <- PRIME::poolReplicates(
    ctss,
    replicates = paste(colData(ctss)$CellLine,
                       colData(ctss)$Type,
                       sep = "_")
)
ctss_pooled <- CAGEfightR::calcTPM(ctss_pooled)

# Subsample to the minimum library depth for fair comparison
min_depth <- min(colData(ctss_pooled)$totalTags)
ctss_pooled_sub  <- PRIME::subsampleTarget(ctss_pooled, target = min_depth)
ctss_pooled_sub  <- CAGEfightR::calcTPM(ctss_pooled_sub)

# Remove singletons
ctss_clean <- PRIME::rmSingletons(ctss_pooled_sub)
```


## Step 3: Assess complexity and noise
See `vignette("ctss-processing")` and `vignette("noise-estimation")` for full details.

### Assess complexity on the replicated data (before pooling)
```{r complexity}
complexity <- PRIME::calcCTSSComplexity(ctss,
                                        step           = 2e6,
                                        minCTSSsupport = 1,
                                        CTSSunexpressed = 0)
```

```{r complexity-plot}
# Set up the dataframe for plotting
df_complexity <- data.frame(
  totalTags   = complexity$totalTags,
  numberCTSSs = complexity$numberCTSSs,
  sample      = complexity$sample
)

# Remove the missing points
df_complexity <- na.omit(df_complexity)

# Plot
ggplot(df_complexity, aes(x = totalTags,
                          y = numberCTSSs,
                          color = sample,
                          group = sample)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_point(size = 1.2) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Total Tags", y = "Number of CTSSs", color = NULL,
       title = "Library complexity saturation curves") +
  theme_bw()
```

### Estimate genomic background noise
```{r noise}
# Requires mask (blacklist + hotspots + exons) and mappable regions
# See vignette("noise-estimation") for full preparation steps
#noise <- PRIME::estimateNoise(
#    ctss_clean,
#    mask       = mask,
#    mappable   = mappable,
#    strand     = "*",
#    inputAssay = "counts",
#    quantiles  = seq(0, 1, 0.0001)
#)
```

## Step 4: Call divergent loci

### Call divergent loci

```{r divergent-loci}
CTSSs_by_chr <- split(ctss_clean, GenomicRanges::seqnames(ctss_clean))

DLs <- lapply(names(CTSSs_by_chr), function(chr) {
    PRIME::divergentLociTCsSummit(
        ctss         = CTSSs_by_chr[[chr]],
        callingAssay = "counts.noSingletons"
    )
})

# Combine results across chromosomes
DLs <- do.call(c, DLs)
```

## Step 5: Predict regulatory elements with PRIMEmodel

Typically, you can run PRIMEmodel::predict() on replicated CTSS data "without any prior filtering."
However, this depends entirely on the specific downstream analysis steps you want the PRIMEmodel to predict.

As an example continuing from the previous steps, 
the ctss_pooled_sub object will be passed through the model to predict the regions of interest 
between the subsampled pooled replicates of K562 whole-cell CAGE with the singleton removed (ctss_clean) from step 2-3. 

Then, those prediction results will be used by PRIMEmodel::predictFocal()
to call each individual replicate within the predicted regions.

This is a downsized version of the analysis performed with pooled samples on the FANTOM5 dataset. 
Further details can be found in the manuscript.

To ensure the process remains computationally efficient,
only chromosomes 16 and 17 will be selected as an example.

```{r subset-pool}
pool_target_samples <- c("K562_C")
ctss_clean_slt <- ctss_clean[, pool_target_samples]
ctss_clean_slt <- GenomeInfoDb::keepSeqlevels(ctss_clean_slt,
                                              c("chr16", "chr17"),
                                              pruning.mode = "coarse")
```

The full setup for the reticulate Python environment is described in
`vignette("prediction")`. The full PRIMEmodel prediction can be run on this
example data when that environment is available:

```{r primemodel, purl=FALSE}
library(reticulate)
library(PRIMEmodel)
library(GenomicRanges)

region_result <- PRIMEmodel::predict(
    ctss_clean_slt,
    python_path     = reticulate::py_config()$python,
    score_threshold = 0.75,
    score_diff      = 0.1,
    num_cores       = NULL,
    keep_tmp        = FALSE
)
```

For a lightweight, self-contained example, load the precomputed prediction
regions bundled with PRIMEmodel. This also keeps the focal prediction example
independent of the optional genome-wide prediction chunk above.

```{r region-result}
region_result <- readRDS(system.file(
    "extdata",
    "predicted_regions_gr.rds",
    package = "PRIMEmodel"
))
```

Run focal prediction on the predicted region for each K562 whole-cell CAGE:

```{r subset-replicates}
target_samples <- c("K562_C1", "K562_C2")
ctss_slt <- ctss[, target_samples]
ctss_slt <- GenomeInfoDb::keepSeqlevels(ctss_slt,
                                        c("chr16", "chr17"),
                                        pruning.mode = "coarse")
```

```{r focal}
focal_result <- PRIMEmodel::predictFocal(
    ctss_slt,
    region_result,
    python_path = reticulate::py_config()$python
  )
```

## Full reproducibility

The complete analysis code for the associated publication (Einarsson,
Navamajiti, et al. 2026) is available at:
<https://github.com/anderssonlab/nucCAGE_PRIME_paper>

That repository contains step-by-step R Markdown notebooks for all analyses
described in the paper.

## See also

- `vignette("getting-started")` — installation of all tools
- `vignette("preprocessing")` — PRIMEprep in depth
- `vignette("ctss-processing")` — CTSS quantification and QC
- `vignette("divergent-loci")` — divergent loci calling
- `vignette("normalization-batches")` — normalization and subsampling
- `vignette("noise-estimation")` — noise estimation
- `vignette("prediction")` — PRIMEmodel prediction in depth
- [PRIMEprep repository](https://github.com/anderssonlab/PRIMEprep)
- [PRIMEmodel repository](https://github.com/anderssonlab/PRIMEmodel)
- [Paper analysis code](https://github.com/anderssonlab/nucCAGE_PRIME_paper)
