---
title: "Estimating Genomic Background Noise"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating Genomic Background Noise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Accurate noise estimation is essential for distinguishing real transcription
initiation events from random background signal. PRIME estimates genomic
background noise by sampling random uniquely-mappable windows that fall
outside known functional regions (e.g., exons, known active loci, blacklisted
regions) and computing the empirical distribution of CTSS expression levels
within those windows.

Two main functions are provided:

- `PRIME::estimateNoise()` — estimates genome-wide background from random windows
- `PRIME::estimateDivergentNoise()` — estimates background specifically for
  divergent transcription

## Setup

```{r libraries}
library(PRIME)
library(rtracklayer)
library(GenomicRanges)
library(GenomicFeatures)
library(GenomeInfoDb)
```

Assumes a CTSS `RangedSummarizedExperiment` object `ctss` is available;
see `vignette("03-ctss-processing")` for how to build one.

## Prepare mask regions

The mask excludes known functional regions from the random window sampling.
A typical mask combines a genomic blacklist, transcription hotspots, and
flanked exons:

```{r mask}
# ENCODE blacklist regions
blacklist <- rtracklayer::import.bed("ENCFF356LFX.bed")

# Transcription hotspot regions (e.g., extended and merged)
hotspots <- rtracklayer::import.bed("hotspots_merged_ext_300.bed")

# Flanked exons from a TxDb object (e.g., TxDb.Hsapiens.UCSC.hg38.knownGene)
exons <- GenomicRanges::flank(
    unlist(GenomicFeatures::exonsBy(txdb, "tx")),
    width = 50,
    both  = TRUE
)

# Combine and reduce to a single non-overlapping mask
mask <- GenomicRanges::reduce(c(blacklist, hotspots, exons))
```

## Import uniquely mappable regions

Noise estimation is restricted to uniquely mappable genomic positions:

```{r mappable}
map.gr.plus  <- rtracklayer::import.bed("uniquely.mappable.plus.bed")
map.gr.minus <- rtracklayer::import.bed("uniquely.mappable.minus.bed")
mappable     <- GenomicRanges::reduce(c(map.gr.plus, map.gr.minus))
```

## Synchronize sequence information

The mask and mappable regions must share sequence levels and info with the
CTSS object:

```{r seqinfo}
GenomeInfoDb::seqlevels(mask, pruning.mode = "coarse")     <- GenomeInfoDb::seqlevels(ctss)
GenomeInfoDb::seqlevels(mappable, pruning.mode = "coarse") <- GenomeInfoDb::seqlevels(ctss)
GenomeInfoDb::seqinfo(mask)     <- GenomeInfoDb::seqinfo(ctss)
GenomeInfoDb::seqinfo(mappable) <- GenomeInfoDb::seqinfo(ctss)
```

## Estimate genomic background noise

```{r estimate-noise}
noise <- PRIME::estimateNoise(
    ctss,
    mask        = mask,
    mappable    = mappable,
    strand      = "*",
    inputAssay  = "counts",
    quantiles   = seq(0, 1, 0.0001)
)
```

The result is a matrix of CTSS expression quantiles estimated from random
background windows. These quantiles can be used to set noise thresholds for
downstream analyses (e.g., filtering low-expressed CTSSs).

## Estimate divergent background noise

For analyses focusing on divergent transcription, `estimateDivergentNoise()`
estimates background using strand-specific mappable regions:

```{r estimate-div-noise}
div_noise <- PRIME::estimateDivergentNoise(
    ctss,
    mask,
    map.gr.minus,
    map.gr.plus
)
```

## Interpretation

- The noise quantiles represent the empirical distribution of background CTSS
  expression levels in non-functional regions.
- A common threshold is the 99th percentile of the noise distribution: CTSSs
  with expression above this level are unlikely to be background noise.
- This threshold is used in divergent loci calling and regulatory element
  prediction to distinguish signal from noise.

## See also

- `vignette("03-ctss-processing")` — building the CTSS object
- `vignette("04-divergent-loci")` — calling divergent loci
- `vignette("07-prediction")` — predicting regulatory elements with PRIMEmodel
- [Paper analysis code — Noise Estimation](https://github.com/anderssonlab/nucCAGE_PRIME_paper/tree/main/9.Noise)
