---
title: "Normalization and Subsampling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Normalization and Subsampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette describes how PRIME handles normalization across libraries and
subsampling for saturation analyses in CTSS datasets.

The examples use a CTSS `RangedSummarizedExperiment` object bundled with PRIME.
See `vignette("ctss-processing")` for how to build this object from BigWig
files.

## Setup

```{r libraries}
library(PRIME)
library(SummarizedExperiment)
```

```{r load-ctss}
ctss <- readRDS(system.file("extdata", "ctss.rds", package = "PRIME"))
```

## Size factor normalization

PRIME provides two size-factor-based normalization functions:

```{r size-factor}
# Compute size factors (e.g., from total tags or an external method)
size_factors <- colData(ctss)$totalTags

# Normalize raw counts by size factors
ctss_norm <- PRIME::normalizeBySizeFactors(
    ctss,
    sizeFactors  = size_factors,
    inputAssay   = "counts",
    outputAssay  = "normalized"
)

# Normalize to TPM scaled by size factors
ctss_norm <- PRIME::TPMnormalizeBySizeFactors(
    ctss,
    sizeFactors  = size_factors,
    inputAssay   = "counts",
    outputAssay  = "normalizedTPM"
)
```

## Subsampling

Subsampling is used to equalize library depth before comparative analyses and
for sequencing saturation experiments.

### Subsample to an absolute target depth

```{r subsample-target}
target_depth <- min(colData(ctss)$totalTags)
ctss_sub <- PRIME::subsampleTarget(ctss, target = target_depth)
```

### Subsample to a proportion of the library

```{r subsample-prop}
ctss_prop <- PRIME::subsampleProportion(ctss, proportion = 0.5)
```

## Batch-aware workflows

The bundled example data do not contain batch labels, so batch normalization is
not run here. For multi-batch datasets, PRIME provides
`PRIME::calcBatchSupport()` to summarize support across batches and
`PRIME::conditionalNormalize()` to normalize within batches. Use a batch vector
with one entry per sample, for example
`batch = colData(ctss)$Batch`, when calling these functions.

## Export normalized BigWig files

After normalization, export per-sample BigWig files for visualization or
downstream tools:

```{r write-bw, purl=FALSE}
# Create the directory if it doesn't already exist
if (!dir.exists("bw_output")) {
  dir.create("bw_output", recursive = TRUE)
}

PRIME::writeBw(
    ctss,
    replicates = "all",
    dir        = "bw_output/"
)
```

## See also

- `vignette("ctss-processing")` - CTSS quantification and basic QC
- `vignette("noise-estimation")` - estimating genomic background noise
- `vignette("end-to-end-workflow")` - complete pipeline walkthrough
- [Paper analysis code](https://github.com/anderssonlab/nucCAGE_PRIME_paper)
