---
title: "Predicting Regulatory Elements with PRIMEmodel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting Regulatory Elements with PRIMEmodel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

[PRIMEmodel](https://anderssonlab.github.io/PRIMEmodel/) uses a LightGBM model
trained on CAGE data to predict the locations of active regulatory elements
(enhancers and promoters) genome-wide or in user-defined focal regions.

PRIMEmodel takes a CTSS `RangedSummarizedExperiment` as input and outputs BED
files with predicted regulatory element coordinates and scores.

Two prediction modes are available:

- **Genome-wide prediction** (`PRIMEmodel::predict()`) — scans the whole genome
  using a sliding-window approach over tag clusters.
- **Focal prediction** (`PRIMEmodel::predictFocal()`) — restricts prediction to
  user-defined regions of interest (e.g., FANTOM5 enhancers or called divergent
  loci).

## Setup

```{r libraries}
library(PRIMEmodel)
library(reticulate)
```

PRIMEmodel requires a Python environment (>= 3.9) with LightGBM. See the
[installation guide](https://anderssonlab.github.io/PRIMEmodel/articles/installation.html)
for details. Full function documentation is available on the
[PRIMEmodel pkgdown site](https://anderssonlab.github.io/PRIMEmodel/reference/).

## Obtaining CTSS input

PRIMEmodel accepts CTSS data produced by either of two approaches:

### Option A: PRIMEmodel convenience function

```{r option-a}
# bw_dir: directory containing BigWig files from PRIMEprep
# design_matrix: data frame with columns Name, BigWigPlus, BigWigMinus
ctss_rse <- PRIMEmodel::plc_get_ctss_from_bw(bw_dir, design_matrix)
ctss_rse <- GenomeInfoDb::keepStandardChromosomes(ctss_rse, pruning.mode = "coarse")
```

### Option B: Standard CAGEfightR / PRIME workflow

```{r option-b}
library(CAGEfightR)

bw_plus  <- rtracklayer::BigWigFileList(design$bw_plus)
bw_minus <- rtracklayer::BigWigFileList(design$bw_minus)
names(bw_plus) <- names(bw_minus) <- rownames(design)

ctss_rse <- CAGEfightR::quantifyCTSSs(plusStrand = bw_plus,
                                      minusStrand = bw_minus,
                                      design      = design)
ctss_rse <- CAGEfightR::calcTotalTags(ctss_rse)
ctss_rse <- GenomeInfoDb::keepStandardChromosomes(ctss_rse, pruning.mode = "coarse")
```

Option B is useful when you have already performed CTSS QC steps with PRIME
(e.g., replicate pooling, subsampling, singleton removal) before running
PRIMEmodel. See `vignette("03-ctss-processing")` for the full CTSS workflow.

## Genome-wide prediction

`PRIMEmodel::predict()` runs the complete 6-step prediction pipeline:

1. **Extract CTSS** — retrieve CTSS counts from the RSE object
2. **Identify tag clusters (TCs)** — call unidirectional TCs genome-wide
3. **Slide through TCs** — apply a sliding window over each TC
4. **Normalized profiles** — normalize CTSS profiles within windows
5. **Predict probabilities** — apply the LightGBM model
6. **Post-processing** — merge windows, apply score threshold, write BED

```{r genome-wide}
result <- PRIMEmodel::predict(
    ctss_rse        = ctss_rse,
    python_path     = reticulate::py_config()$python,
    score_threshold = 0.75,
    score_diff      = 0.1,
    num_cores       = NULL,   # NULL uses all available cores
    keep_tmp        = FALSE
)
```

The result is a `GRanges` object (and corresponding BED file) with predicted
regulatory element coordinates and prediction scores.

### Running via the bash script

For very large datasets or for reproducibility on HPC clusters, the bash
script interface is recommended:

```bash
# Genome-wide prediction
cd PRIMEmodel/genomewide_prediction
./PRIMEmodel.sh --config bash_config_predict.sh --predict

# Run individual steps
./PRIMEmodel.sh --config bash_config_predict.sh -1 -2 -3 -4 -5 -6
```

The individual step flags (`-1` through `-6`) correspond to the six pipeline
steps described above, allowing re-runs from any intermediate checkpoint.

## Focal prediction

`PRIMEmodel::predictFocal()` restricts prediction to a set of pre-defined
genomic regions (e.g., called divergent loci, candidate enhancers):

```{r focal}
# regions_gr: a GRanges object with regions of interest
result_focal <- PRIMEmodel::predictFocal(
    ctss_rse    = ctss_rse,
    regions     = regions_gr,
    python_path = reticulate::py_config()$python
)
```

### Focal prediction via bash

```bash
cd PRIMEmodel/genomewide_prediction
./PRIMEmodel.sh --config bash_config_predictFocal.sh --predictFocal
```

## Choosing between genome-wide and focal prediction

| Mode | Use case |
|---|---|
| Genome-wide | Discovery of all active regulatory elements in the genome |
| Focal | Scoring/classifying a specific set of candidate regions |

Focal prediction is faster and is appropriate when you already have a set of
candidate regions (e.g., from divergent loci calling or an external database).

## Output

Both `predict()` and `predictFocal()` return a `GRanges` object and write
BED files to the specified output directory. Each predicted element has:

- Genomic coordinates (chr, start, end, strand)
- A prediction score (0–1; higher = more confident regulatory element)
- Element type annotation (enhancer / promoter)

## See also

- `vignette("01-getting-started")` — installing PRIMEmodel
- `vignette("03-ctss-processing")` — building CTSS input via CAGEfightR/PRIME
- `vignette("04-divergent-loci")` — calling divergent loci for focal prediction
- `vignette("08-end-to-end-workflow")` — complete pipeline walkthrough
- [PRIMEmodel website](https://anderssonlab.github.io/PRIMEmodel/)
- [PRIMEmodel reference](https://anderssonlab.github.io/PRIMEmodel/reference/)
- [PRIMEmodel installation guide](https://anderssonlab.github.io/PRIMEmodel/articles/installation.html)
- [Paper analysis code — Genome-wide prediction](https://github.com/anderssonlab/nucCAGE_PRIME_paper/tree/main/8.Genomewide_prediction)
