---
title: "Predicting Regulatory Elements with PRIMEmodel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting Regulatory Elements with PRIMEmodel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

[PRIMEmodel](https://anderssonlab.github.io/PRIMEmodel/) uses a LightGBM model
trained on CAGE data to predict the locations of active regulatory elements
(enhancers and promoters) genome-wide or in user-defined focal regions.

PRIMEmodel takes a CTSS `RangedSummarizedExperiment` as input 
and outputs the `GRangesList` with predicted regulatory element coordinates and scores.

Two prediction modes are available:

- **Genome-wide prediction** (`PRIMEmodel::predict()`) — scans the whole genome
  using a sliding-window approach over tag clusters.
- **Focal prediction** (`PRIMEmodel::predictFocal()`) — restricts prediction to
  user-defined regions of interest (e.g. FANTOM5 enhancers or called divergent
  loci, pooled prediction from PRIMEmodel).

## Setup

Setup a PRIME-Compatible Virtual Environment

```{r virtual-env, purl=FALSE}
txt_file <- file.path(find.package("PRIMEmodel"), "envfile", "environment.txt")
txt_content <- readLines(txt_file)

library(reticulate)

# Install Python locally
reticulate::install_python(version = "3.9.22")

# Create a new virtual environment for running PRIMEmodel
virtualenv_create("PRIMEmodel", python = "3.9.22")

# Activate the environment and install dependencies
use_virtualenv("PRIMEmodel")
virtualenv_install(envname = "PRIMEmodel", packages = txt_content)

# Optional (if not already covered by environment.txt)
virtualenv_install(envname = "PRIMEmodel", packages = "matplotlib")
```

```{r libraries, purl=FALSE}
library(PRIMEmodel)
library(GenomicRanges)

# Activate virtual environment
use_virtualenv("PRIMEmodel", required = TRUE)
```

PRIMEmodel requires a Python environment (>= 3.9) with LightGBM. See the
[installation guide](https://anderssonlab.github.io/PRIMEmodel/articles/installation.html)
for details. Full function documentation is available on the
[PRIMEmodel pkgdown site](https://anderssonlab.github.io/PRIMEmodel/reference/).

## Obtaining CTSS input

```{r path-PRIME}
# design_matrix: data frame with columns Name, BigWigPlus, BigWigMinus
# e.g. from PRIME
dir_design <- system.file("extdata", "design_matrix_first10pct.tsv", package = "PRIME")
design <- read.table(dir_design, header = TRUE, sep = "\t")
rownames(design) <- design$Name

# dir_bw: directory containing BigWig files from PRIMEprep
# e.g. from PRIME
dir_bw <- system.file("extdata", "cage_bw", package = "PRIME")
```

PRIMEmodel accepts CTSS data produced by either of two approaches:
### Option A: PRIMEmodel convenience function

```{r option-a}
ctss <- PRIMEmodel::plc_get_ctss_from_bw(dir_bw, design)
ctss <- GenomeInfoDb::keepStandardChromosomes(ctss, pruning.mode = "coarse")
```

### Option B: Standard CAGEfightR / PRIME workflow.

```{r option-b}
library(CAGEfightR)

bw_plus  <- rtracklayer::BigWigFileList(file.path(dir_bw, design$BigWigPlus))
bw_minus <- rtracklayer::BigWigFileList(file.path(dir_bw, design$BigWigMinus))
names(bw_plus) <- names(bw_minus) <- rownames(design)

ctss <- CAGEfightR::quantifyCTSSs(plusStrand = bw_plus,
                                  minusStrand = bw_minus,
                                  design      = design)
ctss <- CAGEfightR::calcTotalTags(ctss)
ctss <- GenomeInfoDb::keepStandardChromosomes(ctss, pruning.mode = "coarse")
```

Option B is useful when you have already performed CTSS QC steps with PRIME
(e.g., replicate pooling, subsampling, singleton removal) before running
PRIMEmodel. See `vignette("03-ctss-processing")` for the full CTSS workflow.

Due to the computationally heavy nature of these processing pipelines,
we highly recommend running this function on a server.
By default, it utilizes half of the available CPU cores for parallelization.
If parallelization is unavailable, the function will automatically fall back to sequential processing,
which will significantly increase the execution time.

If you are running this example on a personal or local computer,
we highly recommend using the pre-prepared .rds object,
which contains only data for chromosomes 16 and 17 to test the function time-efficiently.

```{r rds-ctss}
ctss <- readRDS(system.file("extdata", "ctss_rse_chr16to17.rds", package = "PRIMEmodel"))
```

## Genome-wide prediction

`PRIMEmodel::predict()` runs the complete 6-step prediction pipeline:

1. **Extract CTSS** — retrieve CTSS counts from the RSE object
2. **Identify tag clusters (TCs)** — call unidirectional TCs genome-wide
3. **Slide through TCs** — apply a sliding window over each TC
4. **Normalized profiles** — normalize CTSS profiles within windows
5. **Predict probabilities** — apply the LightGBM model
6. **Post-processing** — merge windows, apply score threshold, write BED

```{r genome-wide}
result <- PRIMEmodel::predict(
    ctss,
    python_path     = reticulate::py_config()$python,
    score_threshold = 0.75,
    score_diff      = 0.1,
    num_cores       = NULL,   # NULL uses half of available cores (max 25 cores)
    keep_tmp        = FALSE
)
```

The result is a `GRangesList` object with predicted regulatory element coordinates 
and prediction scores.

### Running via the bash script

For very large datasets or for reproducibility on HPC clusters, the bash
script interface is recommended:

```bash
# Genome-wide prediction
cd PRIMEmodel/genomewide_prediction
./PRIMEmodel.sh --config bash_config_predict.sh --predict

# Run individual steps
./PRIMEmodel.sh --config bash_config_predict.sh -1 -2 -3 -4 -5 -6
```

The individual step flags (`-1` through `-6`) correspond to the six pipeline
steps described above, allowing re-runs from any intermediate checkpoint.

## Focal prediction

`PRIMEmodel::predictFocal()` restricts prediction to a set of pre-defined
genomic regions (e.g., called divergent loci, candidate enhancers, pooled prediction from PRIMEmodel)

The `regions` object is a GRanges object containing the regions of interest. 
These should either have a uniform width of 401 bp 
or include a 'thick' column to allow internal resizing/extension within the function.

```{r rds-regions}
# regions: a GRanges object with regions of interest
regions <- readRDS(system.file("extdata", "predicted_regions_gr.rds", package = "PRIMEmodel"))
```

```{r focal}
result_focal <- PRIMEmodel::predictFocal(
    ctss,
    regions,
    python_path = reticulate::py_config()$python
)
```

### Focal prediction via bash

```bash
cd PRIMEmodel/genomewide_prediction
./PRIMEmodel.sh --config bash_config_predictFocal.sh --predictFocal
```

## Choosing between genome-wide and focal prediction

| Mode | Use case |
|---|---|
| Genome-wide | Discovery of all active regulatory elements in the genome |
| Focal | Scoring/classifying a specific set of candidate regions |

Focal prediction is faster and is appropriate when you already have a set of
candidate regions (e.g., from divergent loci calling or an external database).

## Output

Both `predict()` and `predictFocal()` return a `GRanges` object and write
BED files to the specified output directory. Each predicted element has:

- Genomic coordinates (chr, start, end, strand)
- A prediction score (0–1; higher = more confident regulatory element)
- Element type annotation (enhancer / promoter)

## See also

- `vignette("01-getting-started")` — installing PRIMEmodel
- `vignette("03-ctss-processing")` — building CTSS input via CAGEfightR/PRIME
- `vignette("04-divergent-loci")` — calling divergent loci for focal prediction
- `vignette("08-end-to-end-workflow")` — complete pipeline walkthrough
- [PRIMEmodel website](https://anderssonlab.github.io/PRIMEmodel/)
- [PRIMEmodel reference](https://anderssonlab.github.io/PRIMEmodel/reference/)
- [PRIMEmodel installation guide](https://anderssonlab.github.io/PRIMEmodel/articles/installation.html)
- [Paper analysis code — Genome-wide prediction](https://github.com/anderssonlab/nucCAGE_PRIME_paper/tree/main/8.Genomewide_prediction)
