---
title: "Installing PRIMEmodel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing PRIMEmodel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide explains how to set up the required **R** and **Python** environments to use the `PRIMEmodel` R package.

Clone the PRIMEmodel repository:

```bash
git clone https://github.com/anderssonlab/PRIMEmodel.git
```

---

## macOS note: libomp required for LightGBM

For macOS users: `libomp` is required for LightGBM to enable OpenMP (multithreading). Without `libomp`, LightGBM may fail to use multithreading properly and can produce silent errors or crashes during training without clear messages. Xcode Command Line Tools and Homebrew are required.

Check whether `libomp` is already installed:

```bash
# Apple Silicon (M1/M2/M3/...)
ls /opt/homebrew/opt/libomp/lib/libomp.dylib

# Intel macOS (x86_64)
ls /usr/local/opt/libomp/lib/libomp.dylib
```

If it is not present, install it:

```bash
brew install libomp
```

On macOS, R (via homebrew libomp) and Python environments (via conda or virtualenv libomp) can conflict when used together through `reticulate`, potentially causing crashes. If this occurs, managing environment variables or aligning libomp paths may be necessary.

---

## R installation for PRIMEmodel

The R package `PRIMEmodel` depends on a mix of CRAN and Bioconductor packages, and a custom GitHub version of `PRIME`.

We recommend using **R 4.4 or higher**. While R versions >= 4.2 can also be used, they may require additional setup steps. For example, on macOS with R 4.2.x, you may need to install system libraries:

```bash
# Core build tools for R packages
brew install gcc pkg-config

# Libraries for graphics, fonts, and rendering
brew install freetype harfbuzz fribidi libpng cairo

# Libraries for image support
brew install jpeg libpng libtiff

# Version control and Git support
brew install libgit2
```

Optional (recommended for macOS users on R 4.2.x): install
[XQuartz](https://www.xquartz.org/) to avoid X11-related warnings and enable
`x11()` graphics.

### Full R setup

Start an R session:

```bash
R
```

**Step 1.** Install required CRAN packages:

```{r install-cran}
install.packages(c(
  "R.utils",
  "assertthat",
  "data.table",
  "future",
  "future.apply",
  "future.callr",
  "foreach",
  "argparse",
  "doParallel",
  "reticulate",
  "arrow",
  "stringr",
  "magrittr"
))
```

**Step 2.** Install BiocManager (if not already installed):

```{r install-biocmanager}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
```

**Step 3.** Install devtools (if not already installed):

```{r install-devtools}
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")
```

**Step 4.** Install CAGEfightR and PRIME:

If errors occur from prerequisite packages, see `vignette("installation")` for
the full PRIME installation guide.

```{r install-prime}
BiocManager::install("CAGEfightR")
devtools::install_github("anderssonlab/PRIME")
```

**Step 5.** Install additional Bioconductor packages not installed with PRIME:

```{r install-bioc-extra}
BiocManager::install("sparseMatrixStats")
```

**Step 6.** Install PRIMEmodel:

```{r install-primemodel}
devtools::install_github("anderssonlab/PRIMEmodel")
```

Alternatively, install from a local `.tar.gz` (the model will be set up in the
PRIME installation directory):

```{r install-primemodel-local}
install.packages("/PATH/TO/PRIMEmodel/PRIMEmodel_1.0.tar.gz")
```

The published PRIMEmodel model can be found at:
<https://doi.org/10.5281/zenodo.17142494>

---

## Python environment setup

PRIMEmodel requires a Python environment (>= 3.9) with LightGBM and supporting
packages. You can prepare this environment in several ways; choose the option
that best fits your workflow.

---

### Option 1: Use an existing Python installation (pip)

If you already have a working Python installation and want to use it directly:

```bash
cd PRIMEmodel

# Install the required packages
pip3 install -r inst/envfile/environment.txt

# Check the path to Python — use this path in PRIMEmodel functions
which python3
```

Verify the installation in R:

```{r verify-pip}
library(GenomicRanges)
library(PRIMEmodel)

plc_focal_example <- PRIMEmodel::predictFocalExample(python_path = "/PATH/TO/YOUR/PYTHON")
plc_example       <- PRIMEmodel::predictExample(python_path = "/PATH/TO/YOUR/PYTHON")
```

---

### Option 2: Virtualenv via `reticulate` in R

This is the recommended method for R-focused workflows that do not need external
software. Note that `use_virtualenv()` must be called at the start of each R
session, and this approach is not ideal for command-line (CLI) use.

```bash
cd PRIMEmodel
R
```

```{r setup-reticulate-venv}
library(reticulate)

# Define the environment name
env_name <- "prime-env"

# Create the virtual environment if it does not exist
virtualenv_create(envname = env_name)

# Use and configure it
use_virtualenv(env_name, required = TRUE)

# Install Python packages
required_pkgs <- readLines(system.file("envfile", "environment.txt", package = "PRIMEmodel"))
reticulate::py_install(packages = required_pkgs, envname = env_name, method = "virtualenv")

# Confirm active Python path
py_config()$python
```

Verify the installation in R:

```{r verify-reticulate-venv}
library(GenomicRanges)
library(PRIMEmodel)

plc_focal_example <- PRIMEmodel::predictFocalExample(python_path = py_config()$python)
plc_example       <- PRIMEmodel::predictExample(python_path = py_config()$python)
```

---

### Option 3: Conda

An `environment.yml` file is included in the `inst/envfile` folder of the
PRIMEmodel repository.

```bash
cd PRIMEmodel

conda env create -f inst/envfile/environment.yml

# Activate the environment
conda activate prime-env

which python3
# Copy this path to use as python_path in R

conda deactivate
```

Verify the installation in R:

```{r verify-conda}
library(GenomicRanges)
library(PRIMEmodel)

plc_focal_example <- PRIMEmodel::predictFocalExample(python_path = "~/.conda/envs/prime-env/bin/python3")
plc_example       <- PRIMEmodel::predictExample(python_path = "~/.conda/envs/prime-env/bin/python3")
```

---

### Option 4: Manual virtualenv (shell)

This is an advanced option for managing the environment outside of R, useful
when working from the shell or with external tools.

```bash
cd PRIMEmodel

# Set up a Python virtual environment
python3 -m venv ~/prime_env
source ~/prime_env/bin/activate
pip3 install -r inst/envfile/environment.txt

which python3
# Copy this path to use as python_path in R

deactivate
```

Verify the installation in R:

```{r verify-manual-venv}
library(GenomicRanges)
library(PRIMEmodel)

plc_focal_example <- PRIMEmodel::predictFocalExample(python_path = "~/prime_env/bin/python3")
plc_example       <- PRIMEmodel::predictExample(python_path = "~/prime_env/bin/python3")
```

---

## Tips

- If you do not have full control over your R and Python environments, errors may
  occur when they try to communicate. We recommend setting up a dedicated virtual
  environment such as a conda environment.
- Always check the current Python path with `which python3` **after** activating
  your environment.
- Use that full path in the `python_path` argument in R.

---

## See also

- `vignette("01-getting-started")` — overview of the PRIME toolkit
- `vignette("installation")` — installing PRIME (R package)
- `vignette("07-prediction")` — genome-wide regulatory element prediction with PRIMEmodel
- [PRIMEmodel repository](https://github.com/anderssonlab/PRIMEmodel)
