---
title: "Tag cluster decomposition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tag cluster decomposition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

A tag cluster (TC) can span tens to hundreds of base pairs and may contain
several distinct transcription initiation peaks — each corresponding to an
individual core promoter. PRIME provides two strategies to **decompose** a
broad TC into finer sub-clusters:

| Method | Function | When to use |
|--------|----------|-------------|
| **Shape-based** | `PRIME::decompose()` with `fn = PRIME::local_maxima_decompose` | Single-sample or pooled data; fast; no replicates needed |
| **Correlation-based** | `PRIME::decomposeCorr()` with `fn = PRIME::corr_decompose` | Multi-sample data; leverages sample-to-sample variation to detect distinct regulatory elements |

Both functions return a `GRanges` object of decomposed sub-clusters annotated
with the same CAGEfightR-style statistics as ordinary tag clusters (`score`,
`thick`, `tpm.dominant_ctss`, etc.), so they slot directly into downstream
CAGEfightR/PRIME workflows.

## Packages

```{r libraries}
library(CAGEfightR)
library(PRIME)
library(GenomicRanges)
library(SummarizedExperiment)
```

## Inputs: CTSSs and tag clusters

The starting point is a `RangedSummarizedExperiment` of CTSSs (from
`CAGEfightR::quantifyCTSSs()`) with pooled signal computed, plus a set of
tag clusters produced by `CAGEfightR::clusterUnidirectionally()`.

### Quantify CTSSs and compute pooled signal

```{r ctss}
ctss <- CAGEfightR::quantifyCTSSs(
  plusStrand  = bw_plus,
  minusStrand = bw_minus,
  design      = design
)
ctss <- CAGEfightR::calcTotalTags(ctss)
ctss <- CAGEfightR::calcTPM(ctss)
ctss <- CAGEfightR::calcPooled(ctss)
```

`calcPooled()` adds a `score` column to the CTSS row-ranges. This pooled
signal is what `PRIME::decompose()` uses as the shape profile for
decomposition — it is required even if you later run the
correlation-based method.

### Call unidirectional tag clusters

```{r tcs}
TCs <- CAGEfightR::clusterUnidirectionally(
  object       = ctss,
  pooledCutoff = 0,
  mergeDist    = 60
)
TCs <- CAGEfightR::quantifyClusters(ctss, TCs)
TCs <- CAGEfightR::calcTPM(TCs)
```

### Optional: restrict to promoter-proximal clusters

If you have a `TxDb` annotation object, you can filter for promoter-associated
clusters before decomposing:

```{r promoter-filter}
# txdb <- AnnotationDbi::loadDb("path/to/annotation.sqlite")
TCs <- CAGEfightR::assignTxType(TCs, txModels = txdb)
TCs_prom <- subset(TCs, txType %in% c("promoter", "proximal", "fiveUTR"))

# Keep only robustly supported clusters
TCs_prom <- CAGEfightR::subsetBySupport(
  TCs_prom,
  inputAssay  = "counts",
  unexpressed = 10,
  minSamples  = 10
)
```

## Shape-based decomposition

`PRIME::decompose()` identifies local maxima in the pooled CTSS profile of
each TC and assigns each base pair to the nearest local maximum according
to a coverage-fraction criterion.

```{r shape-decompose}
decomposed_shape <- PRIME::decompose(
  object = rowRanges(TCs_prom),
  pooled = ctss,
  fn     = PRIME::local_maxima_decompose,
  fraction   = 0.1,
  maximaDist = 20,
  maxGap     = 20,
  mergeDist  = -1,
  smoothPad  = 0
)
```

The `object` argument must be a `GRanges`; pass `rowRanges(TCs_prom)` when
`TCs_prom` is a `RangedSummarizedExperiment`, or pass a bare `GRanges`
directly. `pooled` is the full CTSS `RangedSummarizedExperiment` (must have
`seqlengths` matching `object`).

### Parameter guidance for `local_maxima_decompose`

| Parameter | Default | Effect |
|-----------|---------|--------|
| `fraction` | `0.1` | Minimum coverage relative to a local maximum to be included. Increase (e.g. `0.2`) for more conservative, narrower sub-clusters. |
| `maximaDist` | `20` | Window (bp) used to identify local maxima via a running maximum. Increase for broader promoters; decrease for sharper, better-resolved peaks. |
| `maxGap` | `maximaDist` | Maximum gap (bp) allowed within a sub-cluster. Increase to bridge gapped coverage. |
| `mergeDist` | `-1` | Sub-clusters within this distance are merged. `-1` disables final merging. |
| `smoothPad` | `0` | Extends each sub-cluster by including non-zero CTSSs within this many bp. Useful for noisy signal. |

## Correlation-based decomposition

`PRIME::decomposeCorr()` uses Bayesian change-point analysis on the first
eigenvector of the per-base Pearson correlation matrix across samples. This
leverages differences in expression patterns between samples to identify
boundaries between co-regulated initiation sites.

This method requires multiple samples (replicates or conditions) and the
assay values to be in `ctss` (e.g. `"TPM"`). It is more computationally
intensive than the shape-based approach.

```{r corr-decompose}
decomposed_corr <- PRIME::decomposeCorr(
  object = rowRanges(TCs_prom),
  ctss   = ctss,
  fn     = PRIME::corr_decompose,
  assay  = "TPM",
  thres  = 0.25,
  merge  = TRUE,
  scale  = TRUE
)
```

### Parameter guidance for `corr_decompose`

| Parameter | Default | Effect |
|-----------|---------|--------|
| `assay` | `"TPM"` | Assay used for correlation. `"TPM"` is recommended; raw `"counts"` can be used but may be noisier. |
| `thres` | `0.25` | Posterior probability threshold for Bayesian change-point detection. Lower values detect more change points (more splits); higher values are more conservative. |
| `merge` | `TRUE` | After splitting, merge adjacent sub-clusters whose cross-boundary correlation is positive. Reduces over-splitting. |
| `scale` | `TRUE` | Scale assay values (z-score) before correlation. Recommended to remove library-size effects. |

## Re-quantify decomposed clusters

After decomposition, re-quantify the resulting sub-clusters to get expression
estimates:

```{r requantify}
decomposed_expr <- CAGEfightR::quantifyClusters(ctss, decomposed_shape)
decomposed_expr <- CAGEfightR::calcTPM(decomposed_expr)
```

The output `decomposed_expr` is a `RangedSummarizedExperiment` with the same
structure as a regular tag-cluster RSE, and can be used in all downstream
PRIME/CAGEfightR functions.

## QC: comparing original and decomposed clusters

### Width distributions

```{r qc-widths}
# Width of original TCs
summary(width(rowRanges(TCs_prom)))

# Width of decomposed sub-clusters
summary(width(decomposed_shape))
```

Decomposed sub-clusters should be narrower on average than the parent TCs.
If widths are similar, try decreasing `fraction` or `maximaDist`.

### Number of decomposed clusters vs. original

```{r qc-counts}
n_original   <- length(rowRanges(TCs_prom))
n_decomposed <- length(decomposed_shape)

cat("Original TCs:   ", n_original, "\n")
cat("Decomposed TCs: ", n_decomposed, "\n")
cat("Ratio:          ", round(n_decomposed / n_original, 2), "\n")
```

A ratio close to 1.0 suggests most TCs were not split (possibly good if the
clusters are already narrow). Ratios of 1.2–2.0 are common for broad promoter
regions.

### Width histogram

```{r qc-histogram}
library(ggplot2)

df <- data.frame(
  width  = c(width(rowRanges(TCs_prom)), width(decomposed_shape)),
  source = rep(c("Original TCs", "Decomposed TCs"),
               c(length(rowRanges(TCs_prom)), length(decomposed_shape)))
)

ggplot(df, aes(x = width, fill = source)) +
  geom_histogram(binwidth = 10, position = "identity", alpha = 0.6) +
  scale_x_continuous(limits = c(0, 500)) +
  labs(x = "Width (bp)", y = "Count", fill = NULL,
       title = "TC width before and after decomposition") +
  theme_bw()
```

## Save results

```{r save}
save(
  ctss,
  TCs,
  TCs_prom,
  decomposed_shape,
  decomposed_expr,
  file = "tag_cluster_decomposition.RData"
)
```

## See also

- `vignette("ctss-processing")` — CTSS quantification and tag clustering
- `vignette("divergent-loci")` — divergent transcription analysis
- `vignette("normalization-batches")` — normalization across samples
- [Paper analysis code](https://github.com/anderssonlab/nucCAGE_PRIME_paper)
